Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-09-08 paper/manuscript.tex
Responses to reviewer comments on uncertainty quantification changes by @apbhati1

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method. Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified. This is followed by ensuring identical charges for all MCSS atoms. The approach ensures that the final MCSS is both structurally and chemically identical, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.

@@

\paragraph{Ring breaking and forming.} \label{subsec: Ring breaking and forming}
Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate,bhati2017} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.

@@

The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation. This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling. It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria. This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far. If more samples are needed, the simulations can be resumed, or, started again in different initial conditions enlarging the overall simulation ensemble. 
Bhati \textit{et al.} recommended such a scheme for RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}. They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially and suggest pre-defined criteria (uncertainties below 0.5 kcal/mol and overlap between forward and reverse work values in case of NEQ calculations) to decide on the need of additional calculations (in the way suggested) until these criteria are met.

@@

\subsection{Decorrelating samples for analysis}
\label{sec:decorrelating-samples}
\paragraph{Computing the statistical inefficiency}
The readers should note here that replicas within ensemble simulations are independently run (with randomly assigned starting velocities even when the starting structure is the same) and hence generate decorrelated samples by default. Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022} and hence ensemble simulations are strongly recommended. However, if only a single simulation is available, the procedures below can provide a lower-bound uncertainty estimate.

@@

Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates. Hitherto, MD practitioners have been applying the ergodic theorem that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations. However, it is often ignored that this limit is on the order of a Poincarè recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers. In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be "ergodic" but it must also be ``mixing''~\cite{coveney2016, MDbook2025}. Rate processes and equilibration times are a property of the distribution function, not of individual trajectories. Consequently, neighbouring trajectories in the underlying phase space diverge exponentially which means that a single trajectory can never capture the true behaviour of a system. Rather, one must perform true ensemble averaging through running sufficiently large number of replicas, termed ``ensemble simulations''. This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}. The uncertainty is calculated as the standard error of mean for samples obtained from each replica in the ensemble. The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}. Further, ensemble simulations are also necessary in case of enhanced sampling methods~\cite{bhati2018} as well as microsecond long simulations~\cite{bhati2023}. It is the influence of these reports that journals such as \textit{JCIM} and \textit{Communications Biology} have now mandated the use of ensemble simulations in computational studies~\cite{wan2023}.

@@

The TIES method~\cite{bhati2017} employs a different scheme to obtain the overall variance of the integral. It assumes that the integrand is a Gaussian random process and hence the principle of stochastic integration must be employed to propagate the uncertainties in this integral. The following equation is used to compute the overall variance of the integral combining variance at each $\lambda$ state. $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}\Delta\lambda_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.

>>>>

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method. Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified to ensure chemical similarity between all atom pairs in the final MCSS across both end-points. This is followed by ensuring identical charges for all MCSS atoms by averaging between the two end-points. The resultant change in the sum of atomic charges for all MCSS atoms is uniformly distributed across all atoms in the corresponding alchemical region conserving the overall ligand charge for both end-points. The approach ensures that the final MCSS is both structurally and chemically identical within thresholds, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.

@@

\paragraph{Ring breaking and forming.} \label{subsec: Ring breaking and forming}
Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
TIES also overcomes this issue by using dual topology and ensuring that MCSS atoms are structurally and chemically identical~\cite{bhati2017}.

@@

The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation. This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling. It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019, cournia2017relative, gapsys2021accurate, baumann2021, caves1998}.
A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria. This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far. If more samples are needed, the simulations can be resumed, or, started again in different initial conditions enlarging the overall simulation ensemble. 
Bhati \textit{et al.} recommended such a scheme for RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}. They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially and suggest pre-defined criteria (uncertainties below 0.5 kcal/mol and overlap between forward and reverse work values in case of NEQ calculations) to decide on the need of additional calculations (in the way suggested) until these criteria are met.

@@

\subsection{Decorrelating samples for analysis}
\label{sec:decorrelating-samples}
\paragraph{Computing the statistical inefficiency}
Replicas in ensemble simulations are initialized with velocities drawn independently from the Maxwell–Boltzmann distribution and hence generate decorrelated samples by default. Ensemble approaches are therefore strongly preferred for uncertainty estimation, as single simulations remain constrained by the timescales they can access and may therefore underestimate the true uncertainty. Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022}. When only one simulation is available, the procedures described below provide a lower-bound estimate of the uncertainty.

@@

Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates. Hitherto, MD practitioners have been assuming the validity of the ergodic hypothesis that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations. However, it is often ignored that this limit is on the order of a Poincaré recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers. In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be "ergodic" but it must also be ``mixing''~\cite{coveney2016, MDbook2025}. Rate processes and equilibration times are a property of the distribution function, not of individual trajectories. Consequently, neighbouring trajectories in the underlying phase space diverge exponentially which means that a single trajectory can never capture the true behaviour of a system. Rather, one must perform true ensemble averaging through running sufficiently large number of replicas, termed ``ensemble simulations''. This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}. The uncertainty is calculated as the standard error of the mean for estimates obtained from each replica in the ensemble. The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}. Further, ensemble simulations are also necessary in case of enhanced sampling methods~\cite{bhati2018} as well as microsecond long simulations~\cite{bhati2023}. It is the influence of these reports that journals such as \textit{JCIM} and \textit{Communications Biology} have now mandated the use of ensemble simulations in computational studies~\cite{wan2023}.

@@

The TIES method~\cite{bhati2017} employs a different scheme to obtain the overall variance of the integral. It assumes that the integrand is a Gaussian random process and hence the thermodynamic integral can be viewed as a so-called stochastic integral, which forms the basis of uncertainty propagation in this integral. The following equation is used to compute the overall variance of the integral combining variance at each $\lambda$ state. $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}\Delta\lambda_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.
